The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-28 manuscript/rebuttal.Rmd

### R3 Point 4

> 4. Page 3, Lines 45-47: What about tasks that cannot be automated? I see that this topic is addressed
later. But this part implies that everything can be automated.

RESPONSE: The section has been amended to acknowledge that not all steps can be automated.

> To ensure reproducibility of bioinformatics workflows, they need to be formalized in code wherever
> possible, from inspecting the raw data to generating the outputs that form the conclusions of the
study.

@@

### R3 Point 12

> 12. Page 6, line 32: The master script idea was already mentioned earlier.

RESPONSE: True. We have amended the "Documentation" section to remove the unnecessary reiteration:

> As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
> minimizing the number of commands required.

@@

### R3 Point 18

> 18. Page 10, lines 51-52: I disagree that the risk is small.
> There are many instances of using genomic data to identify individuals who have committed crimes.

RESPONSE: In light of this, we have changed the section in question:

> While deidentification of data has been widely used to safeguard participant privacy, this may not be
> suitable for genomics data due to the availability of existing public genetic data sets that heighten
> reidentification risks [@Heeney2011-ri].
> For example, the surnames of some male participants can be inferred based on the Y chromosome
> variants seen in deidentified genomic data together with public geneology web sites [@Gymrek2013-fz].
> To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
> Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
> frameworks designed to protect personal data in a way that preserves reproducibility [@Rehm2021-kt;@Thorogood2021-kt].

@@

We do not believe any changes to Figure 2 or 4 are necessary.

### R3 Point 20

> 20. Figure 5: I don't think it's really necessary to resummarize the FAIR principles.
> People can go to the source article for that.

RESPONSE: We acknowledge that it can sometimes be tiresome to reiterate things like the FAIR principles.
However summarizing the key points succinctly will be in the interests of readers with limited time.
We note that the original description of the FAIR principles by Wilkinson and colleagues is ~5,000 words,
while this entire manuscript is ~7,000. We note that other reviewers did not insist on removing this
section.

### 2023-08-28 manuscript/5pillars.Rmd
minor spelling

To ensure reproducibility of bioinformatics workflows, they need to be formalised in code wherever
possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
Automated processes remove the need for manual steps, which are time-consuming and
prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or webtools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko,
leading to problems like inadvertent conversion of gene names to dates in databases and supplementary
Excel files [@Zeeberg2004-js; @Ziemann2016-pv; @Abeysooriya2021-uz].
Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.

@@

While deidentification of data has been widely used to safeguard participant privacy, this may not
be suitable for genomics data due to the availability of existing public genetic data sets that heighten
reidentification risks @Heeney2011-ri.
For example, the surnames of some male participants can be inferred based on the Y chromosome variants
seen in deidentified genomic data together with public geneology web sites @Gymrek2013-fz.
To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
frameworks designed to protect personal data in a way that perserves reproducibility
[@Rehm2021-kt;@Thorogood2021-kt].

@@

As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
minimising the number of commands required.
This also makes the `README` documentation much simpler.
For example, the pathway analysis workflow we previously developed could be reproduced and inspected
with just six commands, starting from a new Ubuntu Linux installation [@Ziemann2023-nm].
By minimizing the complexity of reproduction and writing the instructions clearly in the `README`,
reproducibility is made possible to a larger segment of the computational research community,
especially those who are not experts in the languages used.

@@

Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].
This is a direct result of research assessment policies that incentivize journal-level prestige
metrics and number of publications rather than rigor [@Munafo2017-gf; @Oransky2023-zb].
A change in research assessment ("eg: track-record") that recognises and credits contributions toward
reproducibility and rigor would help decrease funds wasted on sloppy research.

>>>>

To ensure reproducibility of bioinformatics workflows, they need to be formalized in code wherever
possible, from inspecting the raw data to generating the outputs that form the conclusions of the study.
Automated processes remove the need for manual steps, which are time-consuming and
prone to errors.
Without an end-to-end automated process, most of the other best practices described here are not
possible.
Scripted workflows, although not always free of errors, enable better auditing and easier reproduction
which would be difficult for graphical tools like spreadsheets or webtools.
Indeed, spreadsheets are prone to data entry, manipulation and formula errors @Panko1996-ko,
leading to problems like inadvertent conversion of gene names to dates in databases and supplementary
Excel files [@Zeeberg2004-js; @Ziemann2016-pv; @Abeysooriya2021-uz].
Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.

@@

While deidentification of data has been widely used to safeguard participant privacy, this may not
be suitable for genomics data due to the availability of existing public genetic data sets that heighten
reidentification risks @Heeney2011-ri.
For example, the surnames of some male participants can be inferred based on the Y chromosome variants
seen in deidentified genomic data together with public geneology web sites @Gymrek2013-fz.
To foster the responsible reuse of sensitive genomic and health data, the Global Alliance for
Genomics and Health (GA4GH) initiative has proposed strategies, technical standards and policy
frameworks designed to protect personal data in a way that preserves reproducibility
[@Rehm2021-kt;@Thorogood2021-kt].

@@

As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
minimizing the number of commands required.
This also makes the `README` documentation much simpler.
For example, the pathway analysis workflow we previously developed could be reproduced and inspected
with just six commands, starting from a new Ubuntu Linux installation [@Ziemann2023-nm].
By minimizing the complexity of reproduction and writing the instructions clearly in the `README`,
reproducibility is made possible to a larger segment of the computational research community,
especially those who are not experts in the languages used.

@@

Another hindrance to reproducibility more broadly is the fixation on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].
This is a direct result of research assessment policies that incentivize journal-level prestige
metrics and number of publications rather than rigor [@Munafo2017-gf; @Oransky2023-zb].
A change in research assessment ("eg: track-record") that recognizes and credits contributions toward
reproducibility and rigor would help decrease funds wasted on sloppy research.
